Gaps in stellar streams as a result of globular cluster fly-bys

### 2024-12-04 paper/main.tex
more changes from sergey

\abstract
  {Thin stellar streams, such as those resulting from the tidal disruption of globular clusters, have long been known and used as probes of the gravitational potential of our Galaxy, both its visible and dark contents. In particular, the presence of under-density regions, or gaps, along these streams is commonly interpreted as being due to the close passage of dark matter sub-haloes. }
 % aims heading (mandatory)
  {In this work, we investigate the perturbations induced on streams by the passage of dense stellar systems, such as globular clusters themselves, to test the possibility that they too may cause the formation of gaps. In particular, we focus on the study of the stream of Palomar 5, a well-known globular cluster in the Galactic halo, which has particularly long tidal tails.  }
 % methods heading (mandatory)
  {For this purpose, we used a particle-test code to simulate the dynamic evolution of Palomar~5 and its tidal tails when Pal~5 is subject to the gravitational action of the Galaxy and its whole system of globular clusters.}
 % results heading (mandatory)
  {Our study shows that the tails of Pal~5 are strongly perturbed by the close passage of other clusters, in particular of NGC~2808, NGC~7078, NGC~104, and that these perturbations induce the formation of gaps in the tails.}
 % conclusions heading (optional), leave it empty if necessary
  {This result suggests that baryonic dense stellar systems such as globular clusters are capable of causing gaps in streams, as already demonstrated in the case of galaxy bar-induced perturbations and giant molecular clouds, and reinforce the idea that the effect of these components should be fully addressed when interpreting the morphology and density structures of observed stellar streams.}

@@

    For the galactic potential, we used the second model from \citet{2017A&A...598A..66P}, a superposition of a thin disk, thick disk, and dark matter halo, with masses and scale lengths provided in Table~1 of \citet{2023A&A...673A..44F}. This model is time-independent throughout our simulations. This model satisfies a series of observational constraints such as local solar stellar density, the galactic rotation curve, etc, similarly to other galactic models such as \texttt{MWpotential2014} from \citet{2015ApJS..216...29B} and \texttt{McMillian2017} from \citet{2017MNRAS.465...76M}. However, To balance data volume and computation time, we use only one galactic potential, which should suffice. \citet{2021MNRAS.505.5978V} found that only a few outer globular clusters are strongly affected by different potential models, with kinematic uncertainties being the dominant factor in orbital solution differences. Similarly, \citet{2024MNRAS.528.5189G} generated a globular cluster mass-loss catalog using seven different potential models and found their debris distributions to be model-independent, similarly to \citet{2023A&A...673A..44F}. While the clusters' positions may depend on the model, we assert that interaction rates and stream formation are unaffected by the choice of galactic model.

    \subsubsection*{ \texttt{Full} simulations}
    There is a primary methodological departure from \citet{2023A&A...673A..44F}.  In that paper, globular clusters evolved under the gravitational effect of the Galaxy only, while now we include also the effect of all the other Galactic globular clusters, by taking into account the direct N-body interactions among them. To quantify these interactions, all clusters are represented by Plummer spheres, each with mass and half-mass radius as reported in the Baumgardt catalogue. In the following of this paper, we refer to these simulations, where the globular clusters interactions are taken into account as \texttt{full} simulations. \\
    For these simulations, we proceed into two steps:
    \begin{enumerate}
      \item First, starting from the Galactocentric positions and velocities of all 165 Galactic globular clusters, we integrate their orbits back in time for 5~Gyr under the influence of the Galaxy itself, and their mutual influence. In the backward integration, the equations of motion for the globular cluster system are thus: 
      \begin{equation}
        \ddot{\vec{r}}_i = -\nabla \Phi(R_i,z_i) + \left.\sum_{j\neq i} \frac{Gm_j}{\left(|\vec{r}_j - \vec{r}_i|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j - \vec{r}_i\right),
      \end{equation}\label{eq:GCNBody} where $\vec{r}$ indicates the galactocentric position, the index $i$ indicates the globular cluster of interest with $R_i,z_i$ being its galactocentric positions in cylindrical coordinates; the index $j$ indicates the other 164 globular clusters that are summed over, where $m_j$ is the mass of the j-th cluster in the sample, $b_j$ is its Plummer scale radius, and $\vec{r_j}$ is its galactocentric position. $\Phi$ represents the Galactic smooth potential \citep[][Model~II, in the present case]{2017A&A...598A..66P}. Note that the masses and sizes of the globular clusters are kept constant in these simulations and not allowed to vary with time. This implies that  we do not take into account in these models the internal evolution of the globular clusters, nor their response to the Galactic tidal field.

      \item Once the positions and velocities of the entire globular cluster system, as to 5~Gyr ago, are found, we sample Palomar~5 with  100,000 particles which follow a Plummer distribution of mass and half-mass radius as in the Baumgardt catalogue.  We then integrate the evolution of these particles forward in time to the present day, by taking into account that each particle feels the gravitational potential of the Galaxy, that of its host cluster, and that of all the other clusters in the Galaxy. Note that we do not take into account self-gravity among particles. The particles expereince the gravitational field yet do not contribute to it. The equation of motion of a generic particle among the 100,000 which populate Pal~5 is thus: 
    \begin{equation}
      \ddot{\vec{r}}_p = -\nabla \Phi(R_p,z_p) + \left.\sum_{j} \frac{Gm_j}{\left(|\vec{r}_j(t) - \vec{r}_p|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j(t)- \vec{r}_p\right),
      \end{equation} \label{eq:equation_of_motion_particle} where the index $p$ represents one of the 100,000 particles of interest, and $j$ indexes over the globular clusters as in Eq.~\ref{eq:GCNBody}

@@

    Regarding the shape of the density distribution, the central peak corresponds to the still-intact globular cluster whose stars have not yet escaped. The density peaks in the stream are of the same order as the cluster itself. This is inconsistent with reality, the cluster should be much more populated than the stream's peak density. Of course, this discrepency a result of our modeling choices. We use the present day mass and radius for Palomar~5 for the whole duration of the simulation. A consequence of this is that the system is modeled as being less dense than it should have been. In turn, our simulations a strong initial mass which adds to the amplitudes of the profile density peaks of Fig.~\ref{fig:stream_on_sky}. This inaccuracy is acceptable for the scope of this work. First is that Palomar~5's tails indeed have more mass than the cluster itself. \citet{2017ApJ...842..120I} reports that there could be three times as much mass in Palomar 5's tails as the cluster itself. Secondly, the exact form of the density distribution is less important than haing a population of particles present that can probe a cluster flyby event.

>>>>

\abstract
%  menntion that only one galactic model should be sufficient to avoid the criticism
  {Thin stellar streams, such as those resulting from the tidal disruption of globular clusters, have long been known and used as probes of the gravitational potential of our Galaxy, both its visible and dark contents. In particular, the presence of under-density regions, or gaps, along these streams is commonly interpreted as being due to the close passage of dark matter sub-haloes. }
 % aims heading (mandatory)
  {In this work, we investigate the perturbations induced on streams by the passage of dense stellar systems, such as globular clusters themselves, to test the possibility that they too may cause the formation of gaps. In particular, we focus on the study of the stream of Palomar 5, a well-known globular cluster in the Galactic halo, which has particularly long tidal tails.  }
 % methods heading (mandatory)
  {For this purpose, we used a particle-test code to simulate the dynamic evolution of Palomar~5 and its tidal tails when Pal~5 is subject to the gravitational action of the Galaxy and its whole system of globular clusters.}
 % results heading (mandatory)
  {Our study shows that the tails of Pal~5 are strongly perturbed by the close passage of other clusters, in particular of NGC~2808, NGC~7078, NGC~104, and that these perturbations induce the formation of gaps in the tails.}
 % conclusions heading (optional), leave it empty if necessary
  {This result suggests that baryonic dense stellar systems such as globular clusters are capable of causing gaps in streams, as already demonstrated in the case of galaxy bar-induced perturbations and giant molecular clouds, and reinforce the idea that the effect of these components should be fully addressed when interpreting the morphology and density structures of observed stellar streams.}

@@

    For the galactic potential, we used the second model from \citet{2017A&A...598A..66P}, a superposition of a thin disk, thick disk, and dark matter halo, with masses and scale lengths provided in Table~1 of \citet{2023A&A...673A..44F}. This model is time-independent throughout our simulations. This model satisfies a series of observational constraints such as local solar stellar density, the galactic rotation curve, etc, similarly to other galactic models such as \texttt{MWpotential2014} from \citet{2015ApJS..216...29B} and \texttt{McMillian2017} from \citet{2017MNRAS.465...76M}. However, To balance data volume and computation time, we use only one galactic potential, which should suffice. \citet{2021MNRAS.505.5978V} found that only a few outer globular clusters are strongly affected by different potential models, with kinematic uncertainties being the dominant factor in orbital solution differences. Similarly, \citet{2024MNRAS.528.5189G} generated a globular cluster mass-loss catalog using seven different potential models and found their debris distributions to be model-independent, similarly to \citet{2023A&A...673A..44F}. While the clusters' positions may depend on the model, we assert that interaction rates and stream formation are unaffected by the choice of galactic potential model.

    \subsubsection*{ \texttt{Full} simulations}
    There is a primary methodological departure from \citet{2023A&A...673A..44F}.  In that paper, globular clusters evolved under the gravitational effect of the Galaxy only, while now we include also the effect of all the other Galactic globular clusters, by taking into account the direct N-body interactions among them. To quantify these interactions, all clusters are represented by Plummer spheres, each with mass and half-mass radius as reported in the Baumgardt catalogue \citep{2021MNRAS.505.5957B}. In the following of this paper, we refer to these simulations, where the globular clusters interactions are taken into account as \texttt{full} simulations. \\
    For these simulations, we proceed into two steps:
    \begin{enumerate}
      \item First, starting from the Galactocentric positions and velocities of all 165 Galactic globular clusters, we integrate their orbits back in time for 5~Gyr under the influence of the Galaxy itself, and their mutual influence. In the backward integration, the equations of motion for the globular cluster system are thus: 
      \begin{equation}
        \ddot{\vec{r}}_i = -\nabla \Phi(R_i,z_i) + \left.\sum_{j\neq i}^{N_{GC}} \frac{Gm_j}{\left(|\vec{r}_j - \vec{r}_i|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j - \vec{r}_i\right),
      \end{equation}\label{eq:GCNBody} where $\vec{r}$ indicates the galactocentric position vector, the index $i$ indicates the globular cluster of interest with $R_i,z_i$ being its galactocentric positions in cylindrical coordinates; the index $j$ indicates the other globular clusters that are summed over. $N_{GC}$ is the total number of globular clusters which in this study is 165, $m_j$ is the mass of the j-th cluster in the sample, $b_j$ is its Plummer scale radius, and $\vec{r_j}$ is its galactocentric position. $\Phi$ represents the Galactic smooth potential \citep[][Model~II, in the present case]{2017A&A...598A..66P}. Note that the masses and sizes of the globular clusters are kept constant in these simulations and not allowed to vary with time. This implies that we do not take into account in these models the internal evolution of the globular clusters, nor their response to the Galactic tidal field.

      \item Once the positions and velocities of the entire globular cluster system, as to 5~Gyr ago, are found, we sample Palomar~5 with  100,000 particles which follow a Plummer distribution of mass and half-mass radius as in the Baumgardt catalogue.  We then integrate the evolution of these particles forward in time to the present day, by taking into account that each particle feels the gravitational potential of the Galaxy, that of its host cluster, and that of all the other clusters in the Galaxy. Note that we do not take into account self-gravity among particles. The particles expereince the gravitational field yet do not contribute to it. The equation of motion of a generic particle among the 100,000 which populate Pal~5 is thus: 
    \begin{equation}
      \ddot{\vec{r}}_p = -\nabla \Phi(R_p,z_p) + \left.\sum_{j}^{N_{GC}} \frac{Gm_j}{\left(|\vec{r}_j(t) - \vec{r}_p|^2 + b_j^2\right)^{3/2}}\right. \left(\vec{r}_j(t)- \vec{r}_p\right),
      \end{equation} \label{eq:equation_of_motion_particle} where the index $p$ represents one of the 100,000 particles of interest, $\vec{r_p}$ being its position, and $j$ indexes over the globular clusters as in Eq.~\ref{eq:GCNBody}.

@@

    Regarding the shape of the density distribution, the central peak corresponds to the still-intact globular cluster whose stars have not yet escaped. The density peaks in the stream are of the same order as the cluster itself. This is inconsistent with reality, the cluster should be much more populated than the stream's peak density. Of course, this discrepancy is a result of our modeling choices. We use the present day mass and radius for Palomar~5 for the whole duration of the simulation. A consequence of this is that the system is modeled as being less dense than it should have been. In turn, our simulations a strong initial mass which adds to the amplitudes of the profile density peaks of Fig.~\ref{fig:stream_on_sky}. This inaccuracy is acceptable for the scope of this work. First is that Palomar~5's tails indeed have more mass than the cluster itself. \citet{2017ApJ...842..120I} reports that there could be three times as much mass in Palomar 5's tails as the cluster itself. Secondly, the exact form of the density distribution is less important than having a population of particles present that can probe a cluster flyby event.
